Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-03-24 index.qmd
Updates 3/24 (JR)

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(grid)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(sjPlot)

@@

-   if HR $> 1$ implies a positive association such that
    $100\% \times (HR - 1)$ is the \\% of greater risk for $Z = z+1$
    when compared to $Z = z$.
-   if HR $< 1$ implies a negatie association such that
    $100\% \times (HR -1)$ is the \\% of decreased risk for $Z = z+1$
    when compared to $Z= z$.
-   if HR $=1$ implies no association. [@nahhas2025]

@@

$$
 H(x) = \int_0^x h(u) du 
$$ {#eq-cumulativehr}

@@

example_coxmodel <- coxph(example_surv_object ~ meal.cal + age, data = lung)
```

@@

CPH assumes the relationship between covariates and the outcome is a
linear relationship [@nahhas2025]. For CPH, the outcome for that
relationship is the log of the hazard rate. This assumption is checked
visually using the Martingale residuals for each covariate
[@nahhas2025]. Martingale residuals are a type of residauls used in
survival analysis that show the discrepancy between the observed and the
expected number of events. The equation is $$
\text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}
$$ The resulting residuals are examined visually and if they are linear
and appear to have a slope of zero, the linearity assumption is not
violated [@Amini2015]. If a model is being generated with multiple
covariates, multiple Martingale residual plots will need to be generated
in order to verify this assumption for each covariate.

@@

The major assumption for CPH is the proportional hazards assumption
which states that the ratio of hazard rates for any two subjects must be
constant at all times. This means the effect of the covariates must be
constant over time [@Bustan2018]. To test this assumption, the
hypothesis test below is used along with Schoenfeld partial residuals.

$$
\begin{align*}
H_0&: \text{the predictor's coefficient does vary with time} \\
H_1&: \text{the predictor's coefficient does not vary with time}
\end{align*}
$$

The scaled Schoenfeld partial residuals are calcuated and returns the
$\chi^2$ value, the degrees of freedom, and p-values for each individual
covariate along with a global value. Using an $\alpha = 0.05$ and the
global p-value, the determination of whether to reject or fail to reject
$H_0$ can be made. In addition to the hypothesis test, these Shoenfeld
residuals can be plotted against time for each covariate to visualize
the assumption. When plotting the residuals against time, if the
assumption is not broken it would be expected that the points show
roughly a straight horizontal line to indicate constant hazard over time
[@nahhas2025].

Individual plots for each covariate are generated which provide a visual
interpretation of the p-values generated from `cox.zph()`. If there is
no clear slope in the residuals on the plot, it signifies further
confirmation the proportional hazards assumption is met.

Below is an example plot demonstrating Shoenfeld residuals plotted
against time to test the proportional hazards assumption. While there is
some curvature in the line, the residuals are roughly horizontal
indicating proportional hazard over time for this covariate. In a model
with multiple covariates, each one would need to be visualized to ensure
it does not break this assumption and can be used in the model.

```{example_phtest <- cox.zph(example_coxmodel)}

@@

**Concordance Index**

To evaluate the model, the Cox regression concordance index can be used.
Two observed pairs are said to be concordant if an event that is
expected to occur before another does occur first. For example, if two
engines are selected from the data and the model predicts one of them to
be at a higher risk of failure than the other and the one with the
higher risk fails first, then the pairs are concordant. If the
alternative happens where the lower risk engine experiences failure
first, then the pairs are discordant.

The concordance index is calculated as follows:

$C = \frac{ConcordantPairs}{ConcordantPairs + DiscordantPairs}$

The concordance index will produce a value between 0 and 1 with 1
meaning all pairs are correctly ordered and 0 being the alternative with
no pairs correctly ordered. The higher the concordance index for the
created model, the more accurately it can model survival for the
data.[@Therneau2017]

### Model Visualization and Prediction

There are multiple different methods to visualize data when creating a
Cox proportional hazards model. Survival curves are a popular way to
visualize survival probabilities over time, but there are other
visualizations that are helpful when interpreting the results of the
model.

Cox proportional hazards models can produce an assortment of useful
metrics to interpret the model as described by the equations above. The
model created can be used to estimate risk of the event occurring, rate
of occurrence, survival probability, and time to event.

#### Forest Plot

@@

```{r, echo=FALSE,warning=FALSE}
plot_model(example_coxmodel,
           dot.size = 1,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

#### Kaplan-Meier Survival Curve

After generating the CPH model, a Kaplan-Meier survival curve can be
generated to view survival probabilities against time. These plots are
important when modeling survival data because they clearly demonstrate
the time where the event being modeled for is expected to occur.

Below is an example survival curve showing survival probabilities over
time. The line represents the median survival probability with the red
area around it showing the 95% confidence interval for the values. The
curve can be interpreted as showing the probability of event would be
expected to happen around a time of 300 where the survival probability
is below 50%.[@Kuitunen2021]

ggsurvplot(example_curve,
           data = lung,
           title = "Example Survival Curve",
           xlab = "Time",
           ylab = "Survival Probability",
           legend = "none")

@@

#### Predicting Time Until Event

Time until event can be calculated by modeling survival probability for
future time values. When survival probability drops below 50%, it will
be assumed the event occurred. Predictions for time until event will be
made on the testing data for each individual's maximum time value. A
column will be added to the table showing the expected time until event.

>>>>

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(grid)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(ezcox)
library(sjPlot)

@@

-   if HR $> 1$ implies a positive association such that
    $100\% \times (HR - 1)$ is the \\% of greater risk for $Z = z+1$
    when compared to $Z = z$.
-   if HR $< 1$ implies a negative association such that
    $100\% \times (HR -1)$ is the \\% of decreased risk for $Z = z+1$
    when compared to $Z= z$.
-   if HR $=1$ implies no association. [@nahhas2025]

@@

$$
\begin{align*}
 H(x) = \int_0^x h(u) du 
\end{align*}
$$ {#eq-cumulativehr}

@@

example_coxmodel <- coxph(example_surv_object ~ meal.cal, data = lung)
```

@@

CPH assumes the relationship between covariates and the outcome is a
linear relationship [@nahhas2025]. For CPH, the outcome for that
relationship is the log of the hazard rate. This assumption is checked
visually using the Martingale residuals for each covariate
[@nahhas2025]. Martingale residuals are a type of residuals used in
survival analysis that show the discrepancy between the observed and the
expected number of events. The equation is $$
\begin{align*}
\text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}
\end{align*}
$$

The resulting residuals are examined visually and if they are linear and
appear to have a slope of zero, the linearity assumption is not violated
[@Amini2015]. If a model is being generated with multiple covariates,
multiple Martingale residual plots will need to be generated in order to
verify this assumption for each covariate.

@@

The major assumption for CPH is the proportional hazards (PH) assumption
which states that the ratio of hazard rates for any two subjects must be
constant at all times. This means the effect of the covariates must be
constant over time [@Bustan2018]. To test this assumption, the
hypothesis test below is used: $$
\begin{align*}
H_0&: \text{the predictor's coefficient does vary with time} \\
H_1&: \text{the predictor's coefficient does not vary with time}
\end{align*}
$$

To generate the p-value for the hypothesis test, the scaled Schoenfeld
partial residuals are calculated for each covariate. The Schoenfeld
partial residuals are the difference between the value of the covariate
and the expected value of the covariate at the time of failure
[@klein2005]. To visually confirm the PH assumption, the plot of the
scaled Schoenfeld partial residuals for each covariate should be a
horizontal line [@harrison2021]. The global p–value will provide the
p-value for all covariates in the model which can be compared to an
$\alpha = 0.05$ to determine if the PH assumption has been met for all
covariates [@nahhas2025].

Violations of the PH assumption can be dealt with in several ways. If
the covariate is categorical, one way is to stratify the covariate which
will generate a separate baseline hazard function for each level of the
covariate [@harrison2021]. This removes the PH assumption for that
covariate, though it will still be assumed for all other covariates in
each level of the stratification [@nahhas2025]. If the covariate is
continuous, one way is to add in a function of time based on the shape
of the hazard ratio plot. This replaces the model's time coefficient
with a function of time as it assumes the covariate varies either
linearly or non-linearly with time [@nahhas2025].

Below is an example plot demonstrating scaled Shoenfeld partial
residuals plotted against time to test the proportional hazards
assumption. While there is some curvature in the line, the residuals are
roughly horizontal indicating proportional hazard over time for this
covariate.

```{r, echo = FALSE}
example_phtest <- cox.zph(example_coxmodel)

@@

Below is an example of the global value that can be generated for the
whole model. In comparing it with an $\alpha = 0.05$, the covariate does
not meet the proportional hazards assumption as $p = 0.128 < \alpha$ .
However, individually, meal.cal does meet the assumption as
$p = 0.044 < \alpha$ . As such, age would need to be adapted based on
one of the above methods in order to include it in the model.

```{r, echo=FALSE}
multi <- coxph(example_surv_object ~ meal.cal + age, data = lung)
ph_tibble <- cox.zph(multi) %>%
  pluck("table") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  as_tibble()
kable(ph_tibble, format = "markdown")
```

Two common model visualizations used for CPH is the forest plot and the
Kaplan-Meier survival curve. The Kaplan-Meier curve assists in
visualizing the initial survival model for the data while the forest
plot assists in visualizing the hazard ratio after generating a model.

**Kaplan Meier Survival Curve**

The Kaplan-Meier (KM) estimator is also known as the Product-Limit
estimator and was proposed by Kaplan and Meier in 1958. The estimator
can be defined as

$$
\hat{S}(t) = \begin{cases}
1 \qquad \qquad \qquad \: \text{if } \; t< t_1,\\
\prod_{t_\leq t}[1 - \frac{d_j}{Y_i}], \quad\text{if} \; t_1 \leq t
\end{cases}
$$ {#eq-km} where,

-   $d_i$ is the number of events at time $t_i$
-   $Y_i$ is the number of individuals who are risk at time $t_i$
    [@klein2005]

KM is a non-parametric estimate which means it makes no assumption of
the shape of the base survival function [@nahhas2025] and is considered
to be well defined up until the largest observed study time $t_{max}$
[@klein2005]. The survival curve provides a visualization of the KM
estimator and are important when modeling survival data because it
demonstrates the time where the event being modeled is expected to
occur.

Below is an example survival curve showing survival probabilities over
time. The line represents the median survival probability with the red
area around it showing the 95% confidence interval for the values. The
curve can be interpreted as showing the probability of event would be
expected to happen around a time of 300 where the survival probability
is below 50% [@Kuitunen2021].

ggsurvplot(example_curve,
           data = lung,
           title = "Example Survival Curve",
           xlab = "Time",
           ylab = "Survival Probability",
           legend = "none")
```

**Forest Plots**

@@

```{r, echo=FALSE,warning=FALSE}
forest <- coxph(example_surv_object ~ age + sex, data = lung)

plot_model(forest,
           dot.size = 1,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
show_forest(lung, covariates = c('age'))

@@

Evaluation of the accuracy of the CPH model is done using the
concordance index which is used to measure the amount of agreement
between two variables. Specifically, the value looks at concordant pairs
and disconcordant pairs. A concordant pair is when either $x_i < x_j$
and $y_i < y_j$ or $x_i > x_j$ and $y_i > y_j$. A discordant pair is
when either $x_i$ or $y_i$ are not in the same place such as $x_i < x_j$
but $y_j < y_i$ [@Therneau2017]. In terms of survival, this can be
thought of as examining when the failure event occurs. If two subjects,
A and B, are compared with subject A having a higher risk, the pair
would be considered concordant if subject A does have the failure event
occur before subject B.

The equation for concordance with CPH is

$$
C =\frac{c + \frac{t_x}{2}}{c + d + t_x}
$$ {#eq-concordance}

-   $c$ is the count of pairs that are concordant
-   $d$ is the count of pairs that are discordant
-   $t_x$ are the pairs tied to the predictor $x$ [@Therneau2017]

The concordance index will produce a value between 0 and 1. A value of 1
means all pairs are correctly ordered and a value 0 being the
alternative with no pairs correctly ordered. Any values between 0 and 1
indicate how accurate the CPH model is with the data [@Therneau2017].

### Prediction

CPH can be used to predict the survival probability at a specific time
$S(t|X=x)$, and the hazard ratio for an individual when compared to a
reference individual $\frac{h(t|X=x)}{h(t|X=x_{ref}}$ [@nahhas2025]. If
the resulting probability is greater than or equal to 50%, it is assumed
the event has not occurred. If the resulting probability is less than
50%, it is assumed the event has occurred. It is also possible to
generate a plot of the estimated survival curves for all values of $t$
that are desired.
